Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-30 concl.tex
added sentence to concl about nonstationarity
few minor fixes on results and concl

>>>>

[[To address the shortcomings mentioned in the previous section, we
    are in the process of repeating our own study with other metrics,
    other forecast methods, and time-series data from other systems,
    as described in the following section.]]

[[Add sentence about ``this curve doesn't hold for nonstationary data,''
talk about what effects that has, tie back to the points to the right
of the MASE wall, and opine about how one might know that this was
happening (and what one might do if it did).]]

### 2014-07-30 results.tex
few minor fixes on results and concl

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be used to predict the future course of
the time series.  
% As mentioned above, though, the effectiveness of any prediction
% strategy depends on how well it leverages the information that is
% available in the data: methods that use linear models, for instance,
% may fail when applied to nonlinear processes.  (This is the basis
% for the second conjecture above.)
In the \svdfive results, this match between model and task manifests
quite clearly.  The MASE scores of the \naive ~and \arima predictions
for this system are $20.870 \pm 0.192$ and $2.370 \pm 0.051$,
respectively: that is, 20.87 and 2.37 times worse than a simple random
walk forecast of the same signals\footnote{The \naive ~MASE score is
  large because of the bimodal nature of the distribution of the
  values of the signal, which makes guessing the mean a particularly
  bad strategy.  The same thing is true of the \svdthree signal.}.  As
before, the positions of these points on a WPE vs. MASE
plot---significantly below and to the right of the shaded
region---should suggest to a practitioner that
%  there is more structure in this signal than the corresponding
% method is able to leverage, and that
a different method might do better.  Indeed, for \svdfive, the LMA
method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster of
results that largely within the shaded region on the WPE-MASE plot.
This is consistent with our second finding: the LMA method can capture
and reproduce the way in which the \svdfive system processes
information, but the \naive ~and \arima prediction methods cannot.

@@

The second effect that can skew MASE scores is nonstationarity.  Since
this metric is normalized by the error of a random-walk forecast
\emph{on the training signal}, differences between the test signal and
training signal can create issues.  This is why the MASE values in
Table~\ref{tab:errors} are $\neq 1.00$ for every random-walk forecast
of every time series: the last 10\% of these signals is significantly
different from the first 90\%.  The deviation from 1.00 will depend on
the process (whether it has multiple regimes, what those regimes look
like, and how it switches between them) as well as the experimental
setup (e.g., sensor precision and data length).  Nonetheless, this
effect suggests another potential heuristic for time-series
prediction: if a practitioner calculates a WPE $>$ \alert{(lower
  bound)} for a time series, and a MASE significantly more than 1,
\alert{why not just $MASE \neq 1.0$?}  this could be an indication
that the signal is nonstationary.  We are in the process of exploring
this idea.

Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_best} were determined from a finite set of
methods and data.  [[But we worked hard to make these experiments
    representative and comprehensive: forecast methods from simple to
    sophisticated; time-series data from a system whose behavior spans
    the dynamical behavior space.]]  While we are cautiously optmistic
about its generality, more exploration will be required to firm up
that conclusion.  Our preliminary work along those lines shows that
data from the Henon map \cite{henon}, the Lorenz system \cite{lorenz},
and the SFI A data set \cite{sfi-data} all fall within the one-$sigma$
volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_best} region, as do various nonlinear
transformations of \svdtwo and \svdsix.

Of course, the geometry of the curves and bounds do not hold for other
error metrics.  The positions of the WPE vs. error points---and the
reasons to choose a particular function to fit to them, or a
theoretical bound on the meaning of the results---may change if one
calculates error in a different way.  The results presented here only
hold for the MASE error metric.  Even though this metric has its
weaknesses, we concur with \cite{MASE} that it is an effective way to
compare prediction error across time-series data of different lengths
and different scales.  [[Point to that paper for extensive evaluation
    and comparison wrt other metrics.]]  We are in the process of
repeating our own study with other metrics, other forecast methods,
and time-series data from other systems.  [[Segue to next section.]]

>>>>

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be used to predict the future course of
the time series.  
% As mentioned above, though, the effectiveness of any prediction
% strategy depends on how well it leverages the information that is
% available in the data: methods that use linear models, for instance,
% may fail when applied to nonlinear processes.  (This is the basis
% for the second conjecture above.)
The MASE scores of the \naive ~and \arima predictions for this system
are $20.870 \pm 0.192$ and $2.370 \pm 0.051$, respectively: that is,
20.87 and 2.37 times worse than a simple random walk forecast of the
same signals\footnote{The \naive ~MASE score is large because of the
  bimodal nature of the distribution of the values of the signal,
  which makes guessing the mean a particularly bad strategy.  The same
  thing is true of the \svdthree signal.}.  As before, the positions
of these points on a WPE vs. MASE plot---significantly below and to
the right of the shaded region---should suggest to a practitioner that
%  there is more structure in this signal than the corresponding
% method is able to leverage, and that
a different method might do better.  Indeed, for \svdfive, the LMA
method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster of
results that largely within the shaded region on the WPE-MASE plot.
This is consistent with our second finding: the LMA method can capture
and reproduce the way in which the \svdfive system processes
information, but the \naive ~and \arima prediction methods cannot.

@@

The second effect that can skew MASE scores is nonstationarity.  Since
this metric is normalized by the error of a random-walk forecast
\emph{on the training signal}, differences between the test signal and
training signal can create issues.  This is why the MASE values in
Table~\ref{tab:error} are not identically one for every random-walk
forecast of every time series: the last 10\% of these signals is
significantly different from the first 90\%.  The deviation from 1.00
will depend on the process---whether it has multiple regimes, what
those regimes look like, and how it switches between them---as well as
the experimental setup (e.g., sensor precision and data length).
Nonetheless, this effect suggests another potential heuristic for
time-series prediction: if a practitioner calculates a WPE $>$
\alert{(lower bound)} for a time series, and a MASE significantly more
than 1, \alert{why not just $MASE \neq 1.0$?}  this could be an
indication that the signal is nonstationary.  We are in the process of
exploring this idea.

Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_best} were determined from a finite set of
methods and data.  [[But we worked hard to make these experiments
    representative and comprehensive: forecast methods from simple to
    sophisticated; time-series data from a system whose behavior spans
    the dynamical behavior space.]]  While we are cautiously optmistic
about its generality, more exploration will be required to firm up
that conclusion.  Our preliminary work along those lines shows that
data from the Henon map \cite{henon}, the Lorenz system \cite{lorenz},
and the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo and \svdsix.

Of course, the geometry of the curves and bounds do not hold for other
error metrics.  The positions of the WPE vs. error points---and the
reasons to choose a particular function to fit to them, or a
theoretical bound on the meaning of the results---may change if one
calculates error in a different way.  The results presented here only
hold for the MASE error metric.  Even though this metric has its
weaknesses, we concur with \cite{MASE} that it is an effective way to
compare prediction error across time-series data of different lengths
and different scales.  [[Point to that paper for extensive evaluation
    and comparison wrt other metrics.]]
